Locality and Effect Reflection

### 2025-07-10 paper.tex
Final tweaks

\newcommand{\nat}{\mathbb{N}}
\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\catobj}{C}
\newcommand{\op}{\mathtt{op}}
\newcommand{\catop}{\cat^{\op}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{{#2}^{#1}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflectraw}[1]{\mathop{\reflectname_{#1}}}
\newcommand{\reflect}[3]{\reflectraw{#1}(#2, #3)}
\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
\newcommand{\perform}[5]{\performraw{#1}{#2}(#3, #4. #5)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reifyraw}[1]{\mathop{\reifyname_{#1}}}
\newcommand{\reify}[3]{\reifyraw{#1}(#2.\,#3)}
\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
\newcommand{\outlname}{\mathrm{out}_L}
\newcommand{\outl}[1]{\mathop{\outlname} #1}
\newcommand{\outrname}{\mathrm{out}_R}
\newcommand{\outr}[1]{\mathop{\outrname} #1}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}
\newcommand{\indintro}[4]{#2_{#1}(#3,\, #4)}
\newcommand{\retname}{\mathrm{ret}}
\newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \,\{#3\}}
\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
\newcommand{\seq}{\,;\,}
\newcommand{\fst}{\mathop{\pi_1}}
\newcommand{\snd}{\mathop{\pi_2}}
\newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
\newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\types}{\mathrel{:}}
\newcommand{\gtypes}{\mathrel{:_{\Box}}}
\newcommand{\cempty}{\varepsilon}
\newcommand{\ccons}[2]{#1;\,#2}
\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
\newcommand{\cdiv}[1]{#1/\Box}
\newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
\newcommand{\letv}[3]{\mathop{\mathrm{let}} #1 = #2 \mathop{\mathrm{in}} #3}
\newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
\newcommand{\gunbox}[3]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
\newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
\newcommand{\sbst}[3]{#1[#2\backslash#3]}
\newcommand{\turnv}{\mathrel{\vdash_V}}
\newcommand{\turnc}{\mathrel{\vdash_C}}
\newcommand{\evmap}{\mathop{ev}}
\newcommand{\currymap}{\mathop{\lambda}}
\newcommand{\reducestov}{\rightsquigarrow_V}
\newcommand{\reducestoc}{\rightsquigarrow_C}
\newcommand{\mif}{\mathop{\mathrm{if}}}
\newcommand{\mthen}{\mathop{\mathrm{then}}}
\newcommand{\melse}{\mathop{\mathrm{else}}}
\newcommand{\embed}[1]{\mathopen{\lbag}#1\mathclose{\rbag}}

@@

  We introduce \emph{effect reflection}, an alternative presentation of
  algebraic effects and handlers that provides, for any algebraic
  signature $\Sigma$ and type $\tau$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau)
    \qquad \cong \qquad
    \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}
  where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
  $\glob \tau$ \textbf{--} sometimes called the \emph{freer monad} of
  $\Sigma$, $\yoneda{\Sigma}$ is a type representing how to inject terms
  of $\Sigma$ into the current ambient monad, and $\glob$ is a comonadic
  \emph{global modality} used to track which values do not depend on the
  ambient monad. This isomorphism emerges naturally from a denotational
  semantics for algebraic effects in terms of presheaves over bounded
  Lawvere theories.

@@

\subsection{Locality and effect reflection}
Locality captures the general idea of a value depending on some notion
of ``location''. We can represent it in a type system using a comonadic
\emph{global modality} $\glob$ that tracks which values do not depend on
their location. Our key insight is that the ambient monad can be viewed
as a form of location and controlled using $\glob$.

Based on this insight, we present \emph{effect reflection}, providing,
for any algebraic signature $\Sigma$ and type $\tau$, an isomorphism:
\begin{equation*}
  \ind{\Sigma}(\glob \tau)
  \qquad \cong \qquad
  \yoneda{\Sigma} \rightarrow \glob \tau
\end{equation*}
where $\ind{\Sigma}(\glob \tau)$ is an inductive type representing the
free algebra of $\Sigma$ at $\glob \tau$ \textbf{--} sometimes called
the \emph{freer monad} of $\Sigma$, and $\yoneda{\Sigma}$ is a type
representing how to inject terms of $\Sigma$ into the current ambient
monad.

@@

Using this support, we can provide effect reflection as a simple API.
We represent algebraic signatures as modules with a single type
constructor, providing $\ind{\Sigma}$ as
\lstinline[style=oxcaml]{'a Term(Sigma).t} and $\yoneda{Sigma}$ as
\lstinline[style=oxcaml]{Handler(Sigma).t}.

@@

  module State_eff = Reflection(State)
  open Term(State)

@@

The key idea is to index the semantics by the ambient monad: rather than
interpreting types as sets, we interpret them as functions from ambient
monads to sets. We represent ambient monads using their corresponding
algebraic theories in the framework of bounded Lawvere theories.

@@

Locality tracks whether a value can 'escape' the current
\emph{region}. By default, the current region ends when a function
returns, so local values cannot be returned from functions.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value escapes its region.} @*)
\end{lstlisting}

The \lstinline[style=oxcaml]{exclave} construct allows functions to
return local values.
\begin{lstlisting}[style=oxcaml]
  let return_local x =
    exclave
      (let (o @ local) = Some x in
       o)
\end{lstlisting}
\lstinline[style=oxcaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression
\lstinline[style=oxcaml]{e}. \lstinline[style=oxcaml]{e} is essentially
executed within the region of the caller of the function and any local
values created within it can be safely returned to that caller.

@@

By default, modes are deep: a local \lstinline[style=oxcaml]{string list}
is a local list of local strings. However, the depth that a mode applies
can be controlled via \emph{modalities}. Record fields can be annotated
with \lstinline[style=oxcaml]{global} modality to indicate that the
contents of the field are global even when the surrounding record is
local. For example:
\begin{lstlisting}[style=oxcaml]
  type 'a gbl = { g : 'a @@ global }
\end{lstlisting}
gives a record whose only field has the \lstinline[style=oxcaml]{global}
modality. The contents of the \lstinline[style=oxcaml]{g} field is always
global, for example:
\begin{lstlisting}[style=oxcaml]
  let project_global (r @ local) =
    r.g
\end{lstlisting}
gives a function of type:
\begin{lstlisting}[style=oxcaml]
  val project_global : 'a gbl @ local -> 'a @ global
\end{lstlisting}
Variant constructor arguments can also be annotated with modalities.

@@

For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t} is
the type of mappings from the operations of \lstinline[style=oxcaml]{O}
into the operations of the current ambient monad. Alternatively,
\lstinline[style=oxcaml]{Handler(O).t} can be described as representing
algebraic effect handlers that are handling the current computation --
hence the name.

@@

A key feature of algebraic effects is how easy they are to compose. We
can use state whilst using asynchronous I/O by simply nesting their
handlers:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ch ->
    handle_state i (fun sh ->
      List.iter (fun fl ->
        let s = await ch (File.read fl) in
        let i = int_of_string s in
        set sh (i + get sh)) files;
      get sh))
\end{lstlisting}

This relies on the fact that the \lstinline[style=oxcaml]{handle_state}
uses \lstinline[style=oxcaml]{reify_local} and so is able to operate on a
local computation that closes over the local handler value
\lstinline[style=oxcaml]{ch}.

If we tried to do something similar using our
\lstinline[style=oxcaml]{handle_gen} handler, then we get an error:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ch ->
    handle_gen (fun gh ->
      List.iter (fun fl ->
        let s = await ch (File.read fl) in
        let i = int_of_string s in
        gen gh i) files))
  (*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
                       (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
  (*@ \textcolor{red}{ This value is local and would escape its region.} @*)
\end{lstlisting}

@@

Composing handlers, like \lstinline[style=oxcaml]{handle_gen}, that use
\lstinline[style=oxcaml]{reify} rather than
\lstinline[style=oxcaml]{reify_local} is a fundamentally more involved
operation. Consider the case of generators that are able to use
asynchronous I/O. We are given a computation that performs both
\lstinline[style=oxcaml]{Await} and \lstinline[style=oxcaml]{Gen}
operations and we'd like to turn it into a stream of integers.  We would
like to be able to freely pass this stream around. In particular, we do
not wish the stream to be prevented from escaping the scope of the
current \lstinline[style=oxcaml]{Await} handler.  However forcing parts
of the stream requires performing additional
\lstinline[style=oxcaml]{Await} operations. Whenever we force part of the
stream we'll need to provide a handler for those
\lstinline[style=oxcaml]{Await} operations.

@@

This is because this composition requiring forwarding is much less
common than nesting and it fundamentally requires each use of the
\lstinline[style=oxcaml]{Await} operation to perform work proportional to
the number of handlers it is forwarded through -- often referred to as a
``linear search''. There has been a tendency to treat this linear search
as some form of deficiency of algebraic effect handlers, but it is
simply a classic trade-off between the extra expressivity allowed by an
indirection and the runtime cost of traversing that indirection. With
our effect reflection API we are able to use nesting for the cases that
it supports -- avoiding the cost of the indirection -- and fall back to
forwarding for the more general cases.

@@

Our language is simply typed, featuring all the usual type-formers. It
uses a variation of fine-grained call-by-value: it has separate syntax
and typing judgements for values and computations. The syntax and typing
rules for the base language are shown in Fig.\ref{fig:typing-standard}.
\begin{figure}
\begin{align*}
  \mathrm{types}, \tau ::=\;& \tau \times \tau \mid 1 \mid \tau + \tau
                              \mid \tau \rightarrow \tau \mid \ldots \\
  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \ldots \\
  \mathrm{values}, v :=\;& x \mid (v, v) \mid \fst v \mid \snd v
                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c} \mid \ldots \\
  \mathrm{computations}, c :=\;& \return{v} \mid \letv{x}{c}{c}
                                 \mid \app{v}{v} \mid \\
                            & \case{v}{x}{c}{x}{c} \mid \ldots
\end{align*}
\begin{mathpar}
\inferrule{\\}
{\lbind{\Gamma}{x}{\tau} \turnv x \types \tau}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \\ \Gamma \turnv v_1 \types \tau_1}
{\Gamma \turnv (v_1, v_2) \types \tau_1 \times \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \fst{v} \types \tau_1}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \snd{v} \types \tau_2}
\and
\inferrule{\\}
{\Gamma \turnv () \types 1}
\and
\inferrule{\Gamma \turnv v \types \tau_1}
{\Gamma \turnv \inl{v} \types \tau_1 + \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_2}
{\Gamma \turnv \inr{v} \types \tau_1 + \tau_2}
\and
\inferrule{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2}
{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau}
{\Gamma \turnc \return{v} \types \tau}
\and
\inferrule{\Gamma \turnc c_1 \types \tau_1 \\
           \lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}
{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2 \\
           \Gamma \turnv v_2 \types \tau_1}
{\Gamma \turnc v_1 v_2 \types \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 + \tau_2 \\
           \lbind{\Gamma}{x_1}{\tau_1} \turnc c_1 \types \tau_3 \\
           \lbind{\Gamma}{x_2}{\tau_2} \turnc c_2 \types \tau_3}
{\Gamma \turnc \case{v}{x_1}{c_1}{x_2}{c_2} \types \tau_3}
\end{mathpar}
\caption{Standard syntax and typing rules}
\label{fig:typing-standard}
\end{figure}

@@

Effect reflection requires, for any signature $\Sigma$ that is being
reflected, an inductive type $\ind{\Sigma}$ to represent the free
algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
  monad} of $\Sigma$~\cite{kiselyov2015freer}.

@@

We also allow the sum of two signatures $\Sigma_1 + \Sigma_2$ to be used
as a signature. It has operations:
\begin{align*}
\{ \; &\mathrm{left}(op_{1,1}) \types \tau_{p_{1,1}} \rightarrowtail \tau_{a_{1,1}} \seq
        \ldots \seq
        \mathrm{left}(op_{1, n}) \types \tau_{p_{1, n}} \rightarrowtail \tau_{a_{1, n}} \\
      &\mathrm{right}(op_{2, 1}) \types \tau_{p_{2, 1}} \rightarrowtail \tau_{a_{2,1}} \seq
        \ldots \seq
        \mathrm{right}(op_{2, m}) \types \tau_{p_{2, m}} \rightarrowtail \tau_{a_{2, m}}
        \; \}
\end{align*}
where $op_{1,1} \seq \ldots \seq op_{1,n}$ are the operations of
$\Sigma_1$ and $op_{2,1} \seq \ldots \seq op_{2,m}$ are the operations
of $\Sigma_2$.

For each operation $op \types \tau_p \rightarrowtail \tau_a$,
$\ind{\Sigma}(\tau)$ has a constructor $\indintro{\Sigma}{op}{v_p}{v_k}$
where $v_p \types \tau_p$ and
$v_k \types \tau_a \rightarrow \ind{\Sigma}(\tau)$. $\ind{\Sigma}(\tau)$
also has a constructor $\ret{\Sigma}{v}$ where $v \types \tau$.

>>>>

\newcommand{\nat}{\mathbb{N}}
\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\catobj}{C}
\newcommand{\op}{\mathtt{op}}
\newcommand{\catop}{\cat^{\op}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{{#2}^{#1}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflectraw}[1]{\mathop{\reflectname_{#1}}}
\newcommand{\reflect}[3]{\reflectraw{#1}(#2, #3)}
\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
\newcommand{\perform}[5]{\performraw{#1}{#2}(#3, #4. #5)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reifyraw}[1]{\mathop{\reifyname_{#1}}}
\newcommand{\reify}[3]{\reifyraw{#1}(#2.\,#3)}
\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
\newcommand{\outlname}{\mathrm{out}_L}
\newcommand{\outl}[1]{\mathop{\outlname} #1}
\newcommand{\outrname}{\mathrm{out}_R}
\newcommand{\outr}[1]{\mathop{\outrname} #1}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}
\newcommand{\indintro}[4]{#2_{#1}(#3,\, #4)}
\newcommand{\retname}{\mathrm{ret}}
\newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \,\{#3\}}
\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
\newcommand{\seq}{\,;\,}
\newcommand{\fst}{\mathop{\pi_1}}
\newcommand{\snd}{\mathop{\pi_2}}
\newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
\newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\types}{\mathrel{:}}
\newcommand{\gtypes}{\mathrel{:_{\Box}}}
\newcommand{\cempty}{\varepsilon}
\newcommand{\ccons}[2]{#1;\,#2}
\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
\newcommand{\cdiv}[1]{#1/\Box}
\newcommand{\bdiv}[1]{\cdiv{\mathop{!_{#1}}}}
\newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
\newcommand{\letv}[3]{\mathop{\mathrm{let}} #1 = #2 \mathop{\mathrm{in}} #3}
\newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
\newcommand{\gunbox}[3]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
\newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
\newcommand{\sbst}[3]{#1[#2\backslash#3]}
\newcommand{\turnv}{\mathrel{\vdash_V}}
\newcommand{\turnc}{\mathrel{\vdash_C}}
\newcommand{\evmap}{\mathop{ev}}
\newcommand{\currymap}{\mathop{\lambda}}
\newcommand{\reducestov}{\rightsquigarrow_V}
\newcommand{\reducestoc}{\rightsquigarrow_C}
\newcommand{\mif}{\mathop{\mathrm{if}}}
\newcommand{\mthen}{\mathop{\mathrm{then}}}
\newcommand{\melse}{\mathop{\mathrm{else}}}
\newcommand{\embed}[1]{\mathopen{\lbag}#1\mathclose{\rbag}}

@@

  We introduce \emph{effect reflection}, an alternative presentation of
  algebraic effects and handlers that provides, for any algebraic
  signature $\Sigma$ and type $\tau$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau)
    \qquad \cong \qquad
    \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}
  where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
  $\glob \tau$, $\yoneda{\Sigma}$ is a type representing how to inject
  terms of $\Sigma$ into the current ambient monad, and $\glob$ is a
  comonadic \emph{global modality} used to track which values do not
  depend on the ambient monad. $\ind{\Sigma}$ is sometimes called the
  \emph{freer monad} of $\Sigma$. This isomorphism emerges naturally
  from a denotational semantics for algebraic effects in terms of
  presheaves over bounded Lawvere theories.

@@

\subsection{Locality and effect reflection}
Locality captures the general idea of a value depending on some notion
of ``location''. Our key insight is that the ambient monad can be viewed
as a form of location, allowing us to reuse existing type system
features for tracking locality.

We can represent locality in a type system using a comonadic
\emph{global modality} $\glob$ that tracks which values do \emph{not}
depend on their location. Values of an ordinary type $\tau$ are assumed
to potentially depend on the current location, whereas values of the
global type $\glob \tau$ are known not to.

In our setting being global means not depending on the ambient
monad. For example, $A \rightarrow B$ represents computations that may
use operations from the ambient monad, while $\glob (A \rightarrow B)$
contains only pure computations that ignore the ambient monad.

Based on this relationship between locality and effects, we present
\emph{effect reflection}, providing, for any algebraic signature
$\Sigma$ and type $\tau$, an isomorphism:
\begin{equation*}
  \ind{\Sigma}(\glob \tau)
  \qquad \cong \qquad
  \yoneda{\Sigma} \rightarrow \glob \tau
\end{equation*}
where $\ind{\Sigma}(\glob \tau)$ is an inductive type representing the
free algebra of $\Sigma$ at $\glob \tau$, and $\yoneda{\Sigma}$ is a
type representing how to inject terms of $\Sigma$ into the current
ambient monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad}
of $\Sigma$~\cite{kiselyov2015freer}.

@@

Using this support, we can provide effect reflection as a simple API.
We represent algebraic signatures as modules with a single type
constructor, providing $\ind{\Sigma}$ as
\lstinline[style=oxcaml]{'a Term(Sigma).t} and $\yoneda{\Sigma}$ as
\lstinline[style=oxcaml]{Handler(Sigma).t}.

@@

  module State_eff = Reflection(State)

@@

The key idea is to index the semantics by the ambient monad: rather than
interpreting types as sets, we interpret them as functions from ambient
monads to sets. We represent ambient monads by their corresponding
algebraic theories using bounded Lawvere theories.

@@

Locality tracks whether a value can escape the current
\emph{region}. By default, the current region ends when a function
returns, so local values cannot be returned from functions.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value escapes its region.} @*)
\end{lstlisting}
The \lstinline[style=oxcaml]{exclave} construct allows functions to
return local values.
\begin{lstlisting}[style=oxcaml]
  let return_local x =
    exclave
      (let (o @ local) = Some x in
       o)
\end{lstlisting}
\lstinline[style=oxcaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression
\lstinline[style=oxcaml]{e}. \lstinline[style=oxcaml]{e} is essentially
executed within the region of the caller of the function and any local
values created within it can be safely returned to that caller.

@@

By default, modes are deep: a local
\lstinline[style=oxcaml]{string list} is a local list of local
strings. However, the depth that a mode applies can be controlled via
\emph{modalities}. Record fields can be annotated with the
\lstinline[style=oxcaml]{global} modality to indicate that the contents
of the field are global even when the surrounding record is local. For
example:
\begin{lstlisting}[style=oxcaml]
  type 'a gbl = { g : 'a @@ global }
\end{lstlisting}
gives a record whose only field has the \lstinline[style=oxcaml]{global}
modality. The contents of the \lstinline[style=oxcaml]{g} field is always
global, for example:
\begin{lstlisting}[style=oxcaml]
  let project_global (r @ local) =
    r.g
\end{lstlisting}
gives a function of type:
\begin{lstlisting}[style=oxcaml]
  val project_global : 'a gbl @ local -> 'a @ global
\end{lstlisting}
Variant constructor arguments can also be annotated with modalities.

@@

For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t}
is the type of mappings from the operations of
\lstinline[style=oxcaml]{O} into the operations of the current ambient
monad. Alternatively, \lstinline[style=oxcaml]{Handler(O).t} can be
described as representing algebraic effect handlers that are handling
the current computation.

@@

A key feature of algebraic effects is how easy they are to compose. We
can use state whilst using asynchronous I/O by simply nesting their
handlers:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ah ->
    handle_state i (fun sh ->
      List.iter (fun fl ->
        let s = await ah (File.read fl) in
        let i = int_of_string s in
        set sh (i + get sh)) files;
      get sh))
\end{lstlisting}

This relies on the fact that the \lstinline[style=oxcaml]{handle_state}
uses \lstinline[style=oxcaml]{reify_local} and so is able to operate on a
local computation that closes over the local await handler value
\lstinline[style=oxcaml]{ah}.

If we tried to do something similar using our
\lstinline[style=oxcaml]{handle_gen} handler, then we get an error:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ah ->
    handle_gen (fun gh ->
      List.iter (fun fl ->
        let s = await ah (File.read fl) in
        let i = int_of_string s in
        gen gh i) files))
  (*@ \textcolor{red}{Error: let s = await ah (File.read fl) in} @*)
                       (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
  (*@ \textcolor{red}{ This value is local and would escape its region.} @*)
\end{lstlisting}

@@

Composing handlers, like \lstinline[style=oxcaml]{handle_gen}, that use
\lstinline[style=oxcaml]{reify} rather than
\lstinline[style=oxcaml]{reify_local} is a fundamentally more involved
operation. Consider the case of generators that are able to use
asynchronous I/O. We are given a computation that performs both
\lstinline[style=oxcaml]{Await} and \lstinline[style=oxcaml]{Gen}
operations and we'd like to turn it into a stream of integers.  We would
like to be able to freely pass this stream around. In particular, we do
not wish the stream to be prevented from escaping the scope of the
current \lstinline[style=oxcaml]{Await} handler.  However forcing parts
of the stream requires performing additional
\lstinline[style=oxcaml]{Await} operations, so whenever we force part of
the stream we'll need to provide a handler for those
\lstinline[style=oxcaml]{Await} operations again.

@@

This is because composition requiring forwarding is much less common
than nesting and it fundamentally requires each use of the
\lstinline[style=oxcaml]{Await} operation to perform work proportional
to the number of handlers it is forwarded through -- often referred to
as a ``linear search''. There has been a tendency to treat this linear
search as some form of deficiency of algebraic effect handlers, but it
is simply a classic trade-off between the extra expressivity allowed by
an indirection and the runtime cost of traversing that indirection. With
our effect reflection API we are able to use nesting for the cases that
it supports -- avoiding the cost of the indirection -- and fall back to
forwarding for the more general cases.

@@

Our language is simply typed, featuring all the usual type-formers. It
uses a variation of fine-grained call-by-value: it has separate syntax
and typing judgements for values and computations. The syntax and typing
rules for the base language are shown in Fig.\ref{fig:typing-standard}.
\begin{figure}
\begin{align*}
  \mathrm{types}, \tau ::=\;& \tau \times \tau \mid 1 \mid \tau + \tau
                              \mid \tau \rightarrow \tau \mid \ldots \\
  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \ldots \\
  \mathrm{values}, v :=\;& x \mid (v, v) \mid \fst v \mid \snd v
                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c} \mid \ldots \\
  \mathrm{computations}, c :=\;& \return{v} \mid \letv{x}{c}{c}
                                 \mid \app{v}{v} \mid \\
                            & \case{v}{x}{c}{x}{c} \mid \ldots
\end{align*}
\begin{mathpar}
\inferrule{\\}
{\lbind{\Gamma}{x}{\tau} \turnv x \types \tau}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \\ \Gamma \turnv v_1 \types \tau_1}
{\Gamma \turnv (v_1, v_2) \types \tau_1 \times \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \fst{v} \types \tau_1}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \snd{v} \types \tau_2}
\and
\inferrule{\\}
{\Gamma \turnv () \types 1}
\and
\inferrule{\Gamma \turnv v \types \tau_1}
{\Gamma \turnv \inl{v} \types \tau_1 + \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_2}
{\Gamma \turnv \inr{v} \types \tau_1 + \tau_2}
\and
\inferrule{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2}
{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau}
{\Gamma \turnc \return{v} \types \tau}
\and
\inferrule{\Gamma \turnc c_1 \types \tau_1 \\
           \lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}
{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2 \\
           \Gamma \turnv v_2 \types \tau_1}
{\Gamma \turnc v_1 v_2 \types \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 + \tau_2 \\
           \lbind{\Gamma}{x_1}{\tau_1} \turnc c_1 \types \tau_3 \\
           \lbind{\Gamma}{x_2}{\tau_2} \turnc c_2 \types \tau_3}
{\Gamma \turnc \case{v}{x_1}{c_1}{x_2}{c_2} \types \tau_3}
\end{mathpar}
\caption{Basic syntax and typing rules}
\label{fig:typing-standard}
\end{figure}

@@

Effect reflection requires, for any signature $\Sigma$ that is being
reflected, an inductive type $\ind{\Sigma}$ to represent the free
algebras of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
  monad} of $\Sigma$~\cite{kiselyov2015freer}.

@@

We also allow the sum of two signatures $\Sigma_1 + \Sigma_2$ to be used
as a signature. It has operations:
\begin{align*}
\{ \; &\mathrm{left}(op_{1,1}) \types \tau_{p_{1,1}} \rightarrowtail \tau_{a_{1,1}}
        \seq \ldots \seq
        \mathrm{left}(op_{1, n}) \types \tau_{p_{1, n}} \rightarrowtail \tau_{a_{1, n}}
        \seq \\
      &\mathrm{right}(op_{2, 1}) \types \tau_{p_{2, 1}} \rightarrowtail \tau_{a_{2,1}}
        \seq \ldots \seq
        \mathrm{right}(op_{2, m}) \types \tau_{p_{2, m}} \rightarrowtail \tau_{a_{2, m}}
        \; \}
\end{align*}
where $op_{1,1} \seq \ldots \seq op_{1,n}$ are the operations of
$\Sigma_1$ and $op_{2,1} \seq \ldots \seq op_{2,m}$ are the operations
of $\Sigma_2$.

For each operation $op \types \tau_p \rightarrowtail \tau_a$,
$\ind{\Sigma}(\tau)$ has a constructor $\indintro{\Sigma}{op}{v_p}{v_k}$
where $v_p \types \glob \tau_p$ and
$v_k \types \glob \tau_a \rightarrow \ind{\Sigma}(\tau)$. $\ind{\Sigma}(\tau)$
also has a constructor $\ret{\Sigma}{v}$ where $v \types \tau$.
